---
title: "Data Exploration"
author: "Khanh Do, Joyce Gill"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(corrplot)
require(leaps)
library(patchwork)
url <- "https://raw.githubusercontent.com/joycegill/Advanced-Statistical-Modeling/main/data/cleaned/FINAL_DATA.csv"
df <- read_csv(url, show_col_types = FALSE)
```

## Notes
Please refer to the [codebook (click here)](https://docs.google.com/spreadsheets/d/15nn2SjonRKGXaUnBLJxZs6E04GZDL0sNgCyN5-q8Vig/edit?gid=1289751215#gid=1289751215) for information on the data, especially on understanding what the variables mean.

# Predict TUITION2 and TUITION3 (In-state/Out-of-state Tuition)

### Helper
```{r}
VARS_TO_KEEP <- c(
  "CONTROL", "STATE", "LOCALE", "OBEREG", "ACTCM50", "SATVR50", "SATMT50", 
  "DVADM01", "DVADM04", "STUFACR", "APPLFEEU", "ENRTOT", 
  "RMFRGNCP", "RMINSTTP", "RMOUSTTP", "SLO6", "SLOA", 
  "CLASIZUND20", "CLASIZ2049", "CLASIZOVE50", "PROPBOR", "AVEDEBT", 
  "AGRNT_T", "GBA6RTT", "GBA4RTT", "TRRTTOT", "EFNRALT",
  "EFFTMEN", "EFFTWOM", "EFWHITT", "ADMCON7", "PGGRRTT"
)

prep_data <- function(data, target) {
  data %>%
    filter(!is.na(.data[[target]]), .data[[target]] > 0) %>%
    select(all_of(target), all_of(VARS_TO_KEEP)) %>%
    mutate(across(any_of(c("SLO6", "SLOA")), ~ ifelse(. == "Yes", 1, 0))) %>%
    mutate(NSLO6 = ifelse(SLO6 == 0, 1, 0)) %>%
    mutate(NSLOA = ifelse(SLO6 == 0, 1, 0)) %>%
    mutate(IS_PUBLIC = ifelse(CONTROL == "Public", 1, 0)) %>%
    mutate(TEST_REQ = ifelse(ADMCON7 == "Required to be considered for admission", 1, 0)) %>%
    mutate(NO_TEST_REQ = ifelse(TEST_REQ == 0, 1, 0)) %>%
    select(where(is.numeric)) %>%
    na.omit()
}

get_best_subsets_modern <- function(df_clean, target_var) {
  form <- as.formula(paste(target_var, "~ ."))
  reg_fit <- regsubsets(form, data = df_clean, nvmax = 10, nbest = 1)
  res_sum <- summary(reg_fit)
  
  data.frame(
    NumVars = 1:length(res_sum$adjr2),
    AdjR2   = res_sum$adjr2,
    res_sum$which
  ) %>% arrange(desc(AdjR2))
}
```

## Corr Plot
```{r}
# Generate the correlation matrices
df_in  <- prep_data(df, "TUITION2")
df_out <- prep_data(df, "TUITION3")

corrplot(cor(df_in), title = "In-State (TUITION2)", mar = c(0,0,1,0), tl.cex = 0.8)
corrplot(cor(df_out), title = "Out-of-State (TUITION3)", mar = c(0,0,1,0), tl.cex = 0.8)
```

## Best Subsets
```{r}
# Run the Analysis
best_in  <- get_best_subsets_modern(df_in, "TUITION2")
best_out <- get_best_subsets_modern(df_out, "TUITION3")

# Visual comparison of models
plot(regsubsets(TUITION2 ~ ., data = df_in, nvmax=24), scale="adjr2", main="In-State Best Subsets")
plot(regsubsets(TUITION3 ~ ., data = df_out, nvmax=24), scale="adjr2", main="Out-of-State Best Subsets")

# Display numerical results
head(best_in, 5)
head(best_out, 5)
```
### Interpretation of Results
High Adjusted $R^2$. The top model (high R^2 not too many variables) is:
```{r}
summary(lm(TUITION2 ~ SATVR50 + APPLFEEU + ENRTOT + RMFRGNCP + AGRNT_T + GBA6RTT + EFFTMEN + IS_PUBLIC, data = df_in))
summary(lm(TUITION3 ~ SATVR50 + APPLFEEU + ENRTOT + RMINSTTP + RMOUSTTP + AGRNT_T + EFFTMEN + IS_PUBLIC, data = df_out))
```

---

# Scatter Plots: Tuition vs Key Predictors

Below we plot out-of-state tuition (TUITION3) against each of the best predictors from the subset regression. 

```{r}
# Long-format for faceting: one row per (school, predictor_name, x_value)
key_vars <- c(
  "ACTCM50" = "ACT composite 50th %ile",
  "STUFACR" = "Student-faculty ratio",
  "GBA6RTT" = "6-year graduation rate",
  "ENRTOT" = "Total enrollment",
  "CLASIZOVE50" = "P(classes >50)",
  "PROPBOR" = "Proportion borrowers"
)
```

```{r}
# Out-of-state tuition vs each key predictor (with trend line)
plots_list <- list()
for (i in seq_along(key_vars)) {
  x_var <- names(key_vars)[i]
  x_lab <- unname(key_vars)[i]
  p <- ggplot(df_out, aes(x = .data[[x_var]], y = TUITION3)) +
    geom_point(alpha = 0.5, size = 2) +
    geom_smooth(method = "lm", se = TRUE, color = "red", linewidth = 0.8) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(title = paste0("TUITION3 vs ", x_lab), x = x_lab, y = "Out-of-state tuition") +
    theme_minimal()
  plots_list[[i]] <- p
}
# 2 columns, 3 rows
wrap_plots(plots_list, ncol = 2)
```

```{r}
# In-state tuition vs same key predictors
plots_list_in <- list()
for (i in seq_along(key_vars)) {
  x_var <- names(key_vars)[i]
  x_lab <- unname(key_vars)[i]
  p <- ggplot(df_in, aes(x = .data[[x_var]], y = TUITION2)) +
    geom_point(alpha = 0.5, size = 2) +
    geom_smooth(method = "lm", se = TRUE, color = "blue", linewidth = 0.8) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(title = paste0("TUITION2 vs ", x_lab), x = x_lab, y = "In-state tuition") +
    theme_minimal()
  plots_list_in[[i]] <- p
}
wrap_plots(plots_list_in, ncol = 2)
```

# Holistic Tuition: Overpriced vs Good Value

## In-state tuition (TUITION2)

```{r}
# Build analysis data: same sample as prep_data so results match earlier models.
# Keep UNITID and INSTNM so we can label "top overpriced" and "good value" schools in tables.
dat_in <- df %>%
  filter(!is.na(TUITION2), TUITION2 > 0) %>%
  select(UNITID, INSTNM, TUITION2, all_of(VARS_TO_KEEP)) %>%
  mutate(across(any_of(c("SLO6", "SLOA")), ~ ifelse(. == "Yes", 1, 0))) %>%
  select(UNITID, INSTNM, TUITION2, ACTCM50, STUFACR, ENRTOT, RMINSTTP, RMOUSTTP, SLO6, CLASIZOVE50, PROPBOR, AGRNT_T, GBA6RTT) %>%
  na.omit()

# Fit best 10-predictor model; residual = actual - predicted (positive = charges more than expected).
lm_in <- lm(TUITION2 ~ ACTCM50 + STUFACR + ENRTOT + RMINSTTP + RMOUSTTP + SLO6 + CLASIZOVE50 + PROPBOR + AGRNT_T + GBA6RTT,
            data = select(dat_in, -UNITID, -INSTNM))
dat_in <- dat_in %>%
  mutate(pred = predict(lm_in), resid = TUITION2 - pred) %>%
  mutate(value_tier = case_when(
    resid >= quantile(resid, 0.9) ~ "Overpriced",   # top 10% residual = overpriced
    resid <= quantile(resid, 0.1) ~ "Good value",   # bottom 10% = good value
    TRUE ~ "Fair value"
  ))
```

```{r}
# Points above the 45 degree line: actual > predicted (overpriced). Below: good value.
ggplot(dat_in, aes(x = pred, y = TUITION2)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(color = value_tier), alpha = 0.7, size = 2) +
  scale_color_manual(values = c("Overpriced" = "red", "Good value" = "green", "Fair value" = "gray")) +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = "TUITION2: Actual vs predicted (in-state)", x = "Predicted tuition", y = "Actual tuition", color = "Value tier") +
  theme_minimal()
```

```{r}
# Residuals centered near 0; right tail = overpriced, left tail = good value.
ggplot(dat_in, aes(x = resid)) +
  geom_histogram(bins = 25, fill = "blue", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  labs(title = "TUITION2: Residuals (actual - predicted)", x = "Residual ($)", y = "Count") +
  theme_minimal()
```

```{r}
# Rank by residual: highest = most overpriced, lowest = best value.
top_over_in  <- dat_in %>% arrange(desc(resid)) %>% slice_head(n = 10) %>% select(INSTNM, TUITION2, pred, resid)
top_value_in <- dat_in %>% arrange(resid)       %>% slice_head(n = 10) %>% select(INSTNM, TUITION2, pred, resid)
knitr::kable(top_over_in, caption = "Top 10 overpriced (TUITION2)", digits = 0)
knitr::kable(top_value_in, caption = "Top 10 good value (TUITION2)", digits = 0)
```

```{r}
# Check whether "overpriced" schools have higher graduation rates (outcome by value tier).
ggplot(dat_in, aes(x = value_tier, y = GBA6RTT, fill = value_tier)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Overpriced" = "red", "Good value" = "green", "Fair value" = "gray")) +
  scale_x_discrete(limits = c("Good value", "Fair value", "Overpriced")) +
  labs(title = "TUITION2: 6-year graduation rate by value tier", x = "Value tier", y = "Graduation rate (%)") +
  theme_minimal() + theme(legend.position = "none")
```

```{r}
# Test: does mean graduation rate differ across value tiers? Significant F suggests tiers are not equivalent.
anova_in <- aov(GBA6RTT ~ value_tier, data = dat_in)
summary(anova_in)
```

## Out-of-state tuition (TUITION3)

```{r}
# Same logic as TUITION2: same sample definition, keep INSTNM for tables, use best 10-predictor model for TUITION3.
dat_out_holistic <- df %>%
  filter(!is.na(TUITION3), TUITION3 > 0) %>%
  select(UNITID, INSTNM, TUITION3, all_of(VARS_TO_KEEP)) %>%
  mutate(across(any_of(c("SLO6", "SLOA")), ~ ifelse(. == "Yes", 1, 0))) %>%
  select(UNITID, INSTNM, TUITION3, ACTCM50, DVADM04, STUFACR, APPLFEEU, ENRTOT, RMINSTTP, RMOUSTTP, SLO6, AGRNT_T, GBA6RTT) %>%
  na.omit()

lm_out <- lm(TUITION3 ~ ACTCM50 + DVADM04 + STUFACR + APPLFEEU + ENRTOT + RMINSTTP + RMOUSTTP + SLO6 + AGRNT_T + GBA6RTT,
             data = select(dat_out_holistic, -UNITID, -INSTNM))
dat_out_holistic <- dat_out_holistic %>%
  mutate(pred = predict(lm_out), resid = TUITION3 - pred) %>%
  mutate(value_tier = case_when(
    resid >= quantile(resid, 0.9) ~ "Overpriced",
    resid <= quantile(resid, 0.1) ~ "Good value",
    TRUE ~ "Fair value"
  ))
```

```{r}
# Same as TUITION2: 45° line separates overpriced (above) from good value (below).
ggplot(dat_out_holistic, aes(x = pred, y = TUITION3)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(color = value_tier), alpha = 0.7, size = 2) +
  scale_color_manual(values = c("Overpriced" = "red", "Good value" = "green", "Fair value" = "gray")) +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = "TUITION3: Actual vs predicted (out-of-state)", x = "Predicted tuition", y = "Actual tuition", color = "Value tier") +
  theme_minimal()
```

```{r}
# Residual distribution; interpretation same as TUITION2.
ggplot(dat_out_holistic, aes(x = resid)) +
  geom_histogram(bins = 25, fill = "blue", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  labs(title = "TUITION3: Residuals (actual - predicted)", x = "Residual ($)", y = "Count") +
  theme_minimal()
```

```{r}
# Rank by residual; show top 10 overpriced and top 10 best value.
top_over_out  <- dat_out_holistic %>% arrange(desc(resid)) %>% slice_head(n = 10) %>% select(INSTNM, TUITION3, pred, resid)
top_value_out <- dat_out_holistic %>% arrange(resid)       %>% slice_head(n = 10) %>% select(INSTNM, TUITION3, pred, resid)
knitr::kable(top_over_out, caption = "Top 10 overpriced (TUITION3)", digits = 0)
knitr::kable(top_value_out, caption = "Top 10 good value (TUITION3)", digits = 0)
```

```{r}
# Outcome (graduation rate) by value tier; same idea as TUITION2.
ggplot(dat_out_holistic, aes(x = value_tier, y = GBA6RTT, fill = value_tier)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Overpriced" = "red", "Good value" = "green", "Fair value" = "gray")) +
  scale_x_discrete(limits = c("Good value", "Fair value", "Overpriced")) +
  labs(title = "TUITION3: 6-year graduation rate by value tier", x = "Value tier", y = "Graduation rate (%)") +
  theme_minimal() + theme(legend.position = "none")
```

```{r}
# Test whether mean graduation rate differs by value tier for out-of-state tuition.
anova_out <- aov(GBA6RTT ~ value_tier, data = dat_out_holistic)
summary(anova_out)
```

## Summary
Basically what I did is:

1. Predict in-state and out-of-state tuition based on 10 predictors from the best-subset model. 

2. Residual = actual - predicted. So positive = overpriced schools, and negative = good value schools because they charge less than expected. 

3. I created tiers by saying the top 10% are overpriced and the bottom 10% are underpriced and kept the rest as "fair value".

4. Addded tables that show the names of the top 10 most overpriced and most underpriced schools. 


### Top 40 variables to predict Tuition and Debt
```{r}
og_df <- read_csv(url, show_col_types = FALSE)
cleaned_df <- og_df %>%
  select(
    UNITID,
    INSTNM,
    CONTROL,
    ADMCON7,
    ACTCM50,
    DVADM01,
    DVADM04,
    SATVR50,
    SATMT50,
    PROPBOR,
    AVEDEBT,
    CLASIZUND20,
    CLASIZ2049,
    CLASIZOVE50,
    APPLFEEU,
    TUITPL1,
    TUITPL3,
    RMBRDAMT,
    TUITION2,
    TUITION3,
    COAISGS2,
    CHG2AT3,
    CHG3AT3,
    STUFACR,
    ENRTOT,
    EFUG1ST,
    EFUGTRN,
    RMINSTTP,
    RMOUSTTP,
    RMFRGNCP,
    EFFTMEN,
    EFFTWOM,
    EFWHITT,
    EFNRALT,
    UAGRNTP,
    AGRNT_T,
    UFLOANN,
    UDGPGRNTN,
    GBA6RTT,
    GBA4RTT,
    TRRTTOT
  )

write_csv(cleaned_df, "./data/cleaned/TOP_VARS_PREDICT_TUITION_N_DEBT.csv")
```
