---
title: "Data Exploration"
author: "Khanh Do"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(corrplot)
require(leaps)
library(patchwork)
url <- "https://raw.githubusercontent.com/joycegill/Advanced-Statistical-Modeling/main/data/cleaned/FINAL_DATA.csv"
df <- read_csv(url, show_col_types = FALSE)
```

## Notes
Please refer to the [codebook (click here)](https://docs.google.com/spreadsheets/d/15nn2SjonRKGXaUnBLJxZs6E04GZDL0sNgCyN5-q8Vig/edit?gid=1289751215#gid=1289751215) for information on the data, especially on understanding what the variables mean.

# Predict TUITION2 and TUITION3 (In-state/Out-of-state Tuition)

### Helper
```{r}
VARS_TO_KEEP <- c(
  "CONTROL", "LOCALE", "ACTCM50", "SATVR50", "SATMT50", 
  "DVADM01", "DVADM04", "STUFACR", "APPLFEEU", "ENRTOT", 
  "RMFRGNCP", "RMINSTTP", "RMOUSTTP", "SLO6", "SLOA", 
  "CLASIZUND20", "CLASIZOVE50", "PROPBOR", "AVEDEBT", 
  "AGRNT_T", "GBA6RTT", "GBA4RTT", "TRRTTOT", "EFNRALT"
)

prep_data <- function(data, target) {
  data %>%
    filter(!is.na(.data[[target]]), .data[[target]] > 0) %>%
    select(all_of(target), all_of(VARS_TO_KEEP)) %>%
    mutate(across(any_of(c("SLO6", "SLOA")), ~ ifelse(. == "Yes", 1, 0))) %>%
    select(where(is.numeric)) %>%
    na.omit()
}

get_best_subsets_modern <- function(df_clean, target_var) {
  form <- as.formula(paste(target_var, "~ ."))
  reg_fit <- regsubsets(form, data = df_clean, nvmax = 10, nbest = 1)
  res_sum <- summary(reg_fit)
  
  data.frame(
    NumVars = 1:length(res_sum$adjr2),
    AdjR2   = res_sum$adjr2,
    res_sum$which
  ) %>% arrange(desc(AdjR2))
}
```

## Corr Plot
```{r}
# Generate the correlation matrices
df_in  <- prep_data(df, "TUITION2")
df_out <- prep_data(df, "TUITION3")

corrplot(cor(df_in), title = "In-State (TUITION2)", mar = c(0,0,1,0), tl.cex = 0.8)
corrplot(cor(df_out), title = "Out-of-State (TUITION3)", mar = c(0,0,1,0), tl.cex = 0.8)
```

## Best Subsets
```{r}
# Run the Analysis
best_in  <- get_best_subsets_modern(df_in, "TUITION2")
best_out <- get_best_subsets_modern(df_out, "TUITION3")

# Visual comparison of models
plot(regsubsets(TUITION2 ~ ., data = df_in, nvmax=24), scale="adjr2", main="In-State Best Subsets")
plot(regsubsets(TUITION3 ~ ., data = df_out, nvmax=24), scale="adjr2", main="Out-of-State Best Subsets")

# Display numerical results
head(best_in, 5)
head(best_out, 5)
```
### Interpretation of Results
The 6, 7, 8, 9, and 10 variables-models all have high Adjusted $R^2$. The top model is:
```{r}
summary(lm(TUITION2 ~ ACTCM50 + STUFACR + ENRTOT + RMINSTTP + RMOUSTTP + SLO6 + CLASIZOVE50 + PROPBOR + AGRNT_T + GBA6RTT, data = df_in))
summary(lm(TUITION3 ~ ACTCM50 + DVADM04 + STUFACR + APPLFEEU + ENRTOT + RMINSTTP + RMOUSTTP + SLO6 + AGRNT_T + GBA6RTT, data = df_out))
```

**Consistent Drivers (Vertical Columns)**:

- RMINSTTP (Percent of first-time undergraduates - in-state)

- ENRTOT (Total enrollment)

- AGRNT_T (Total amount of federal, state, local or institutional grant aid awarded to full-time first-time undergraduates)

- STUFACR (Student-faculty ratio)

- GBA6RTT (Graduation rate - Bachelor degree within 6 years, total)

**Key Differences**:

- In-State: More sensitive to CLASIZOVE50 (Big class size) than Out-of-state

- Out-of-State: More sensitive to ACTCM50 (ACT Composite 50th percentile score) and APPLFEEU (Undergraduate program application fee) than In-state