---
title: "Graphs"
author: "Joyce Gill, Khanh Do"
date: "2026-02-10"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidyverse)
library(effectsize)
library(corrplot)
library(patchwork)
```


```{r echo=FALSE}
## Read Data
adm_test_f2024 <- read.csv("data/custom_data/ADM_TEST_F2024.csv")
inst_chars_dir_2024_25 <- read.csv("data/custom_data/INST_CHARS_DIR_2024_25.csv")
fe_f2024 <- read.csv("data/custom_data/FE_F2024.csv")
outcome_cohort_2016_17 <- read.csv("data/custom_data/OUTCOME_COHORT_2016_17.csv")
grad_feq_var_cohort_2018_21 <- read.csv("data/custom_data/GRAD_FEQ_VAR_COHORT_2018_21.csv")
finance_2023_24 <- read.csv("data/custom_data/FINANCE_2023_24.csv")
costs_f2024 <- read.csv("data/custom_data/COSTS_F2024.csv")
fin_aid_2023_24 <- read.csv("data/custom_data/FIN_AID_2023_24.csv")
completion_cip_2023_24 <- read.csv("data/custom_data/COMPLETION_CIP_2023_24.csv")
class_size <- read.csv("data/custom_data/CLASS_SIZE_2024_25.csv")
```
## SAT Scores by Admission Requirement
Do schools that require SAT scores have higher average SAT scores than schools that do not require them?
- Schools that do not require SAT seem to have higher overall SAT scores, and it seems like there are a lot more schools that do not require.
```{r}
# Create SAT requirement variable and calculate total SAT score
adm_test_f2024_prep <- adm_test_f2024 %>%
  mutate(
    unitid = as.character(unitid),
    sat_required = ifelse(
      ADM2024.Admission.test.scores == "Required to be considered for admission",
      "Required", "Not Required"
    ),
    sat_total = ADM2024.SAT.Evidence.Based.Reading.and.Writing.50th.percentile.score + 
                ADM2024.SAT.Math.50th.percentile.score
  )

# Remove missing values and add institution names
adm_test_f2024_clean <- adm_test_f2024_prep %>%
  filter(!is.na(sat_total)) %>%
  left_join(
    inst_chars_dir_2024_25 %>%
      mutate(unitid = as.character(unitid)) %>%
      select(unitid, institution_name = institution.name),
    by = "unitid"
  )

# Create boxplot with individual school points
ggplot(adm_test_f2024_clean, aes(x = sat_required, y = sat_total)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.3, width = 0.2, aes(color = sat_required)) +
  labs(x = "SAT Requirement", y = "Total SAT Score", 
       title = "SAT Scores by Admission Requirement",
       subtitle = "Each point represents a school") +
  theme(legend.position = "none")

# Statistical tests
t.test(sat_total ~ sat_required, data = adm_test_f2024_clean)
cohens_d(sat_total ~ sat_required, data = adm_test_f2024_clean)
```

## SAT Requirement vs Admission Rate
Do schools that require SAT scores have different admission rates compared to schools that do not require them?
- Schools that do not require SAT scores seem to have higher acceptance rates, and it seems like a lot more schools do not require.

```{r}
# Create SAT requirement variable and convert admission rate to numeric
sat_admission_rate <- adm_test_f2024 %>%
  mutate(
    unitid = as.character(unitid),
    sat_required = ifelse(
      ADM2024.Admission.test.scores == "Required to be considered for admission",
      "Required", "Not Required"
    ),
    admit_rate = as.numeric(DRVADM2024.Percent.admitted...total)
  ) %>%
  filter(between(admit_rate, 0, 100)) %>%
  left_join(
    inst_chars_dir_2024_25 %>%
      mutate(unitid = as.character(unitid)) %>%
      select(unitid, institution_name = institution.name),
    by = "unitid"
  )

# Create boxplot with individual school points
ggplot(sat_admission_rate, aes(x = sat_required, y = admit_rate)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.3, width = 0.2, aes(color = sat_required)) +
  labs(
    x = "SAT Requirement",
    y = "Admission Rate (%)",
    title = "Admission Rate by SAT Requirement",
    subtitle = "Each point represents a school"
  ) +
  theme(legend.position = "none")

# Statistical tests
t.test(admit_rate ~ sat_required, data = sat_admission_rate)
cohens_d(admit_rate ~ sat_required, data = sat_admission_rate)
```

## SAT Requirements by State
Are SAT requirements geographically clustered across states?
- Feels like more Southern schools overall, but a little hard to tell.

```{r}
# Merge SAT requirement data with institutional characteristics
sat_state_requirements <- adm_test_f2024 %>%
  mutate(
    unitid = as.character(unitid),
    sat_required = ADM2024.Admission.test.scores == "Required to be considered for admission"
  ) %>%
  select(unitid, sat_required) %>%
  inner_join(
    inst_chars_dir_2024_25 %>%
      mutate(unitid = as.character(unitid)) %>%
      select(unitid, state = HD2024.State.abbreviation, institution_name = institution.name),
    by = "unitid"
  ) %>%
  filter(!is.na(state))

# Calculate SAT requirement rate by state (only states with at least 5 schools)
state_rates <- sat_state_requirements %>%
  group_by(state) %>%
  summarise(rate = mean(sat_required), n = n(), .groups = "drop") %>%
  filter(n >= 5)

# Create bar chart with school count labels
ggplot(state_rates, aes(x = reorder(state, rate), y = rate)) +
  geom_col() +
  geom_text(aes(label = paste0("n=", n)), hjust = -0.1, size = 2.5) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "State",
    y = "Percent Requiring SAT",
    title = "SAT Requirement Rate by State",
    subtitle = "Labels show number of schools per state"
  )
```

## SAT Requirements by State and Sector
Are SAT requirements geographically clustered even after accounting for school type (public vs. private)?
- More publics schools seem to require SAT.

```{r}
# Merge SAT requirement data with state and sector information
sat_state_sector <- adm_test_f2024 %>%
  mutate(
    unitid = as.character(unitid),
    sat_required = ADM2024.Admission.test.scores == "Required to be considered for admission"
  ) %>%
  select(unitid, sat_required) %>%
  inner_join(
    inst_chars_dir_2024_25 %>%
      mutate(unitid = as.character(unitid)) %>%
      select(unitid, state = HD2024.State.abbreviation, sector = HD2024.Sector.of.institution, institution_name = institution.name),
    by = "unitid"
  ) %>%
  filter(
    !is.na(state),
    sector %in% c("Public, 4-year or above", "Private not-for-profit, 4-year or above")
  )

# Calculate SAT requirement rate by state and sector
state_sector_rates <- sat_state_sector %>%
  group_by(state, sector) %>%
  summarise(rate = mean(sat_required), n = n(), .groups = "drop") %>%
  filter(n >= 5)

# Create faceted bar chart with school count labels
ggplot(state_sector_rates, aes(x = reorder(state, rate), y = rate)) +
  geom_col() +
  geom_text(aes(label = paste0("n=", n)), hjust = -0.1, size = 2.5) +
  coord_flip() +
  facet_wrap(~ sector) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "State",
    y = "Percent Requiring SAT",
    title = "SAT Requirement Rates by State and Sector",
    subtitle = "Restricted to 4-year institutions (states with 5 or more schools per sector). Labels show number of schools."
  )
```
## SAT Requirements by Region

```{r}
## SAT Requirements by Region – Raw Counts

# Create SAT requirement + merge in region info
sat_region <- adm_test_f2024 %>%
  mutate(
    unitid = as.character(unitid),
    sat_required = ADM2024.Admission.test.scores ==
      "Required to be considered for admission"
  ) %>%
  select(unitid, sat_required) %>%
  inner_join(
    inst_chars_dir_2024_25 %>%
      mutate(unitid = as.character(unitid)) %>%
      select(
        unitid,
        region = `HD2024.Bureau.of.Economic.Analysis..BEA..regions`
      ),
    by = "unitid"
  ) %>%
  filter(!is.na(region))

# Count schools by region and SAT requirement
region_counts <- sat_region %>%
  mutate(sat_required = ifelse(sat_required, "Required", "Not required")) %>%
  group_by(region, sat_required) %>%
  summarise(n = n(), .groups = "drop")

## Bar Chart
ggplot(region_rates, aes(x = reorder(region, rate), y = rate)) +
  geom_col() +
  geom_text(aes(label = paste0("n=", n)), hjust = -0.1, size = 2.8) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Region",
    y = "Percent Requiring SAT",
    title = "SAT Requirement Rates by Region",
    subtitle = "Labels show total number of schools per region"
  )
```

## Enrollment Growth vs Graduation Rates
Do schools that grow enrollment faster sacrifice graduation rates, or do some manage both?
- 

```{r}
# Calculate first-time enrollment as percentage of total enrollment
enroll_prep <- fe_f2024 %>%
  mutate(
    unitid = as.character(unitid),
    total_enroll = as.numeric(DRVEF2024.Total..enrollment),
    first_time = as.numeric(DRVEF2024.First.time.degree.certificate.seeking.undergraduate.enrollment)
  ) %>%
  filter(!is.na(total_enroll), !is.na(first_time), total_enroll > 0, first_time > 0) %>%
  mutate(growth_pct = 100 * first_time / total_enroll) %>%
  select(unitid, growth_pct)

# Get 6-year graduation rates
grad_prep <- grad_feq_var_cohort_2018_21 %>%
  mutate(unitid = as.character(unitid)) %>%
  mutate(grad_rate = as.numeric(DRVGR2024.Graduation.rate...Bachelor.degree.within.6.years..total)) %>%
  filter(!is.na(grad_rate), between(grad_rate, 0, 100)) %>%
  select(unitid, grad_rate)

# Merge enrollment and graduation data with institution names
enrollment_growth_graduation <- enroll_prep %>%
  inner_join(grad_prep, by = "unitid") %>%
  left_join(
    inst_chars_dir_2024_25 %>%
      mutate(unitid = as.character(unitid)) %>%
      select(unitid, institution_name = institution.name),
    by = "unitid"
  )

# Identify outliers for labeling (top 5% and bottom 5% by graduation rate)
outliers <- enrollment_growth_graduation %>%
  mutate(
    grad_rate_rank = percent_rank(grad_rate),
    growth_rank = percent_rank(growth_pct)
  ) %>%
  filter(grad_rate_rank >= 0.95 | grad_rate_rank <= 0.05 | growth_rank >= 0.95 | growth_rank <= 0.05)

# Create scatter plot with labeled outliers
ggplot(enrollment_growth_graduation, aes(x = growth_pct, y = grad_rate)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  geom_text(data = outliers, aes(label = institution_name), 
            size = 2, hjust = 0, vjust = 0, check_overlap = TRUE, nudge_x = 0.5, nudge_y = 0.5) +
  labs(
    x = "First-Time Enrollment (%)",
    y = "6-Year Graduation Rate (%)",
    title = "Enrollment Growth vs Graduation Rates",
    subtitle = "Labels show schools in top/bottom 5% by graduation rate or growth"
  )
```
## Enrollment Growth vs Graduation Rates by Sector
Is the relationship between enrollment growth and graduation rates consistent across public and private institutions?

```{r}
# Add sector information to enrollment-graduation data
growth_grad_sector <- enrollment_growth_graduation %>%
  inner_join(
    inst_chars_dir_2024_25 %>%
      mutate(unitid = as.character(unitid)) %>%
      select(unitid, sector = HD2024.Sector.of.institution),
    by = "unitid"
  ) %>%
  filter(sector %in% c("Public, 4-year or above", "Private not-for-profit, 4-year or above"))

# Identify outliers for labeling by sector
outliers_sector <- growth_grad_sector %>%
  group_by(sector) %>%
  mutate(
    grad_rate_rank = percent_rank(grad_rate),
    growth_rank = percent_rank(growth_pct)
  ) %>%
  ungroup() %>%
  filter(grad_rate_rank >= 0.95 | grad_rate_rank <= 0.05 | growth_rank >= 0.95 | growth_rank <= 0.05)

# Create faceted scatter plot with labeled outliers
ggplot(growth_grad_sector, aes(x = growth_pct, y = grad_rate)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  geom_text(data = outliers_sector, aes(label = institution_name), 
            size = 2, hjust = 0, vjust = 0, check_overlap = TRUE, nudge_x = 0.5, nudge_y = 0.5) +
  facet_wrap(~ sector) +
  labs(
    x = "First-Time Enrollment (%)",
    y = "6-Year Graduation Rate (%)",
    title = "Enrollment Growth vs Graduation Rate by Sector",
    subtitle = "Labels show schools in top/bottom 5% by graduation rate or growth within each sector"
  )
```

## Nonlinear Relationship: Enrollment Growth vs Graduation Rates
Is there a nonlinear relationship between enrollment growth and graduation rates, suggesting an optimal growth rate?

```{r}
# Identify outliers for labeling
outliers_nonlinear <- enrollment_growth_graduation %>%
  mutate(
    grad_rate_rank = percent_rank(grad_rate),
    growth_rank = percent_rank(growth_pct)
  ) %>%
  filter(grad_rate_rank >= 0.95 | grad_rate_rank <= 0.05 | growth_rank >= 0.95 | growth_rank <= 0.05)

# Create scatter plot with loess smooth to explore nonlinearity
ggplot(enrollment_growth_graduation, aes(x = growth_pct, y = grad_rate)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", span = 0.5) +
  geom_text(data = outliers_nonlinear, aes(label = institution_name), 
            size = 2, hjust = 0, vjust = 0, check_overlap = TRUE, nudge_x = 0.5, nudge_y = 0.5) +
  labs(
    title = "Nonlinear Relationship Between Growth and Graduation Rates",
    x = "First-Time Enrollment (%)",
    y = "6-Year Graduation Rate (%)",
    subtitle = "Labels show schools in top/bottom 5% by graduation rate or growth"
  )

# Fit quadratic model to test for nonlinear relationship
quad_model <- lm(grad_rate ~ growth_pct + I(growth_pct^2), data = enrollment_growth_graduation)
summary(quad_model)
```

## Faculty-to-Student Ratio vs Graduation Rates
Does having more faculty per student always translate to better graduation outcomes?

```{r}
# Get student-faculty ratio data
faculty_data <- fe_f2024 %>%
  mutate(
    unitid = as.character(unitid),
    student_faculty_ratio = as.numeric(EF2024D.Student.to.faculty.ratio)
  ) %>%
  filter(!is.na(student_faculty_ratio), between(student_faculty_ratio, 5, 50)) %>%
  select(unitid, student_faculty_ratio)

# Get graduation rates
grad_data <- grad_feq_var_cohort_2018_21 %>%
  mutate(unitid = as.character(unitid)) %>%
  mutate(grad_rate = as.numeric(DRVGR2024.Graduation.rate...Bachelor.degree.within.6.years..total)) %>%
  filter(!is.na(grad_rate), between(grad_rate, 0, 100)) %>%
  select(unitid, grad_rate)

# Merge data with institution names
faculty_ratio_graduation <- faculty_data %>%
  inner_join(grad_data, by = "unitid") %>%
  left_join(
    inst_chars_dir_2024_25 %>%
      mutate(unitid = as.character(unitid)) %>%
      select(unitid, institution_name = institution.name),
    by = "unitid"
  )

# Identify outliers for labeling (top 5% and bottom 5%)
outliers_faculty <- faculty_ratio_graduation %>%
  mutate(
    grad_rate_rank = percent_rank(grad_rate),
    ratio_rank = percent_rank(student_faculty_ratio)
  ) %>%
  filter(grad_rate_rank >= 0.95 | grad_rate_rank <= 0.05 | ratio_rank >= 0.95 | ratio_rank <= 0.05)

# Create scatter plot with loess smooth
ggplot(faculty_ratio_graduation, aes(x = student_faculty_ratio, y = grad_rate)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess") +
  geom_text(data = outliers_faculty, aes(label = institution_name), 
            size = 2, hjust = 0, vjust = 0, check_overlap = TRUE, nudge_x = 0.5, nudge_y = 0.5) +
  labs(
    x = "Student–Faculty Ratio",
    y = "6-Year Graduation Rate (%)",
    title = "Student–Faculty Ratio vs Graduation Rate",
    subtitle = "Labels show schools in top/bottom 5% by graduation rate or faculty ratio"
  )
```

## Class-size vs Tuition Cost
```{r}
# MERGE THE DATASETS
# We join 'class_size' and 'costs_f2024' on UNITID, making sure IDs are the same type
# UNITID is the unique identifier for all IPEDS institutions.

analysis_df <- class_size %>%
  # Ensure UNITID is character for joining
  mutate(UNITID = as.character(UNITID)) %>%
  # Join with costs, also converting unitid to character
  inner_join(
    costs_f2024 %>%
      mutate(unitid = as.character(unitid)),
    by = c("UNITID" = "unitid")
  ) %>%
  # Join institution names from INST_CHARS
  inner_join(
    inst_chars_dir_2024_25 %>%
      mutate(unitid = as.character(unitid)) %>%
      select(unitid, institution_name = institution.name),
    by = c("UNITID" = "unitid")
  ) %>%
  select(
    UNITID,
    institution_name,
    STUFACR,
    CLASIZUND20,
    CLASIZOVE50,
    `COST1_2024.Out.of.state.average.tuition.for.full.time.undergraduates`
  ) %>%
  filter(
    !is.na(`COST1_2024.Out.of.state.average.tuition.for.full.time.undergraduates`) &
      !is.na(STUFACR)
  )

# Identify outliers for labeling
outliers_class <- analysis_df %>%
  mutate(
    tuition_rank = percent_rank(`COST1_2024.Out.of.state.average.tuition.for.full.time.undergraduates`),
    ratio_rank = percent_rank(STUFACR)
  ) %>%
  filter(tuition_rank >= 0.95 | tuition_rank <= 0.05 | ratio_rank >= 0.95 | ratio_rank <= 0.05)

# Predicting Out of state average tuition for full time undergraduates
# based on Student-to-Faculty Ratio and Class Size
ggplot(analysis_df, aes(x = STUFACR, y = `COST1_2024.Out.of.state.average.tuition.for.full.time.undergraduates`)) +
  geom_point(aes(size = CLASIZUND20), alpha = 0.3, color = "darkblue") +
  geom_smooth(method = "lm", formula = y ~ x, color = "red", se = TRUE) +
  geom_text(data = outliers_class, aes(label = institution_name), 
            size = 2, hjust = 0, vjust = 0, check_overlap = TRUE, nudge_x = 0.5, nudge_y = 500) +
  labs(title = "Predicting 2024-25 Tuition by Student-Faculty Ratio",
       subtitle = "Bubbles sized by probability of classes having <20 students. Labels show outliers.",
       x = "Student-to-Faculty Ratio",
       y = "Out of state average tuition",
       size = "Prob. Small Classes") +
  theme_minimal()

ggplot(analysis_df, aes(x = STUFACR, y = `COST1_2024.Out.of.state.average.tuition.for.full.time.undergraduates`)) +
  geom_point(aes(size = CLASIZOVE50), alpha = 0.3, color = "darkblue") +
  geom_smooth(method = "lm", formula = y ~ x, color = "red", se = TRUE) +
  geom_text(data = outliers_class, aes(label = institution_name), 
            size = 2, hjust = 0, vjust = 0, check_overlap = TRUE, nudge_x = 0.5, nudge_y = 500) +
  labs(title = "Predicting 2024-25 Tuition by Student-Faculty Ratio",
       subtitle = "Bubbles sized by probability of classes having >50 students. Labels show outliers.",
       x = "Student-to-Faculty Ratio",
       y = "Out of state average tuition",
       size = "Prob. Big Classes") +
  theme_minimal()
```


## Average Tuition vs Combined Food and Housing Charge

```{r}
df_costs <- costs_f2024 %>%
  select(
    unitid,
    InState = COST1_2024.In.state.average.tuition.for.full.time.undergraduates,
    OutState = COST1_2024.Out.of.state.average.tuition.for.full.time.undergraduates,
    RoomBoard = COST1_2024.Combined.food.and.housing.charge
  ) %>%
  filter(!is.na(InState) & !is.na(OutState) & !is.na(RoomBoard)) %>%
  # Ensure unitid is character before joining to institutional characteristics
  mutate(unitid = as.character(unitid)) %>%
  left_join(
    inst_chars_dir_2024_25 %>%
      mutate(unitid = as.character(unitid)) %>%
      select(unitid, institution_name = institution.name),
    by = "unitid"
  )

# MULTIPLE LINEAR REGRESSION
# Model: Does tuition (In vs Out) predict housing costs?
cost_model <- lm(RoomBoard ~ InState + OutState, data = df_costs)
summary(cost_model)

# Identify outliers for labeling
outliers_costs <- df_costs %>%
  mutate(
    instate_rank = percent_rank(InState),
    outstate_rank = percent_rank(OutState),
    roomboard_rank = percent_rank(RoomBoard)
  ) %>%
  filter(instate_rank >= 0.95 | instate_rank <= 0.05 | outstate_rank >= 0.95 | outstate_rank <= 0.05 | 
         roomboard_rank >= 0.95 | roomboard_rank <= 0.05)

# 3. VISUALIZATION: Dual-Panel Comparison
# Plot A: In-State Tuition vs Room & Board
p1 <- ggplot(df_costs, aes(x = InState, y = RoomBoard)) +
  geom_point(alpha = 0.3, color = "darkblue") +
  geom_smooth(method = "lm", color = "blue") +
  geom_text(data = outliers_costs, aes(label = institution_name), 
            size = 2, hjust = 0, vjust = 0, check_overlap = TRUE, nudge_x = 500, nudge_y = 500) +
  labs(title = "In-State Tuition vs. Living Costs",
       x = "Avg In-State Tuition", y = "Combined Food/Housing",
       subtitle = "Labels show outliers") +
  theme_minimal()

# Plot B: Out-of-State Tuition vs Room & Board
p2 <- ggplot(df_costs, aes(x = OutState, y = RoomBoard)) +
  geom_point(alpha = 0.3, color = "darkred") +
  geom_smooth(method = "lm", color = "red") +
  geom_text(data = outliers_costs, aes(label = institution_name), 
            size = 2, hjust = 0, vjust = 0, check_overlap = TRUE, nudge_x = 500, nudge_y = 500) +
  labs(title = "Out-of-State Tuition vs. Living Costs",
       x = "Avg Out-of-State Tuition", y = "Combined Food/Housing",
       subtitle = "Labels show outliers") +
  theme_minimal()

p1 + p2 + plot_annotation(title = "Tuition Levels as Predictors of Housing Charges (2024-25)")
```
## Potential Graphs:
Tuition vs Post-Graduation Salary: Are Higher Tuition Levels Associated with Higher Earnings?

Tuition vs Instructional Spending: Are Higher-Tuition Schools Actually Investing More?
