---
title: "Best Subset for Predicting Private Undergraduate Insititution In-State Tuition"
author: "Khanh Do"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(corrplot)
require(leaps)
library(patchwork)
require(mosaic)
require(corrplot)
require(leaps)
library(ggplot2)
library(reshape2)
```

# Read in Data

```{r message = FALSE, warning = FALSE, echo = FALSE}
url <- "https://raw.githubusercontent.com/joycegill/Advanced-Statistical-Modeling/main/data/cleaned/FINAL_DATA.csv"
df <- read_csv(url, show_col_types = FALSE)
```

Read csv from this [link](https://raw.githubusercontent.com/joycegill/Advanced-Statistical-Modeling/main/data/cleaned/FINAL_DATA.csv) into a dataframe `df`.

Code book available [here](https://docs.google.com/spreadsheets/d/15nn2SjonRKGXaUnBLJxZs6E04GZDL0sNgCyN5-q8Vig/edit?gid=1289751215#gid=1289751215).

Each row is unique by column UNITID.

# 1. Data Cleaning

For this analysis, we will focus on only Private school.

```{r echo=FALSE}
favstats(TUITION2 ~ CONTROL, data = df) %>%
  select(CONTROL, min, median, max, sd, n, missing)

cat("Number of Private schools = ", 80 + 778, "\n")
cat("Number of columns = ", ncol(df), "\n")
```

## 1st Cleaning

The next code does the following.

- filter the schools down to only Private for-profit and Private not-for-profit
- create dummy variables to distinguish between for-profit and not-for-profit
- turn any categorical variables into dummy variables (if there are n levels, there will be n-1 extra variables)
- remove the original categorical variables because we will use the created dummies for analysis from now on.

```{r}
private1 <- df %>%
  filter(CONTROL %in% c("Private for-profit", "Private not-for-profit")) %>%
  # categorical variables (if there are n levels, there will be n-1 columns)
  mutate(
    FORPROFIT = ifelse(CONTROL == "Private for-profit", 1, 0),
    ADMCON7_REQUIRED = ifelse(ADMCON7 == "Required to be considered for admission", 1, 0),
    APPFEEUW_YES = ifelse(APPFEEUW == "Yes", 1, 0), # application fee waiver
    TUITPL1_YES = ifelse(TUITPL1 == "Yes", 1, 0),   # tuition guaranteed plan
    TUITPL3_YES = ifelse(TUITPL3 == "Yes", 1, 0),   # tuition payment plan
    ALLONCAM_YES = ifelse(ALLONCAM == "Yes", 1, 0), # Stay on-campus required
    ASSOC1_YES = ifelse(ASSOC1 == "Yes", 1, 0),     # member of NCAA
    BOARD_YES = ifelse(BOARD != "No", 1, 0),        # provide meal plan
    ROOM_YES = ifelse(ROOM == "Yes", 1, 0),         # provide housing
    SEMESTER = ifelse(CALSYS == "Semester", 1, 0),  # calendar system
    HBCU_YES = ifelse(HBCU == "Yes", 1, 0),         # Historically Black 
    HOSPITAL_YES = ifelse(HOSPITAL == "Yes", 1, 0), # has hospital
    OPENPUBL_YES = ifelse(OPENPUBL == "Institution is open to the public", 1, 0), 
    RELAFFIL_NO = ifelse(RELAFFIL == "Not applicable", 1, 0),  # not religious
    SLO6_YES = ifelse(SLO6 == "Yes", 1, 0),         # has study abroad
    SLOA_YES = ifelse(SLOA == "Yes", 1, 0),         # has undergraduate research
    STUSRV2_YES = ifelse(STUSRV2 == "Yes", 1, 0),   # has career service
    STUSRV3_YES = ifelse(STUSRV3 == "Yes", 1, 0),   # work-study to pay tuition
  
    # LOCALE: degree of urbanization
    ## not include Rural
    LO_SUBURB = ifelse(LOCALE %in% c("Suburb: Small", "Suburb: Large", "Suburb: Midsize"), 1, 0),
    LO_TOWN = ifelse(LOCALE %in% c("Town: Distant", "Town: Remote", "Town: Fringe"), 1, 0),
    LO_CITY = ifelse(LOCALE %in% c("City: Midsize", "City: Small", "City: Large"), 1, 0),
    
    # HDEGOFR1: Highest degree offered
    ## not include Bachelor
    HDEOFR_DOC = ifelse(HDEGOFR1 %in% c("Doctor's degree - research/scholarship", "Doctor's degree - professional practice", "Doctor's degree - research/scholarship and professional practice"), 1, 0),
    HDEOFR_MAS = ifelse(HDEGOFR1 %in% c("Master's degree"), 1, 0),
    
    # OBEREG: Bureau of Economic Analysis (BEA) regions
    ## not include U.S. Service schools
    NORTHEAST = ifelse(OBEREG %in% c(
      "New England (CT, ME, MA, NH, RI, VT)", 
      "Mid East (DE, DC, MD, NJ, NY, PA)"), 1, 0),
    MIDWEST = ifelse(OBEREG %in% c(
      "Great Lakes (IL, IN, MI, OH, WI)", 
      "Plains (IA, KS, MN, MO, NE, ND, SD)"), 1, 0),
    SOUTH = ifelse(OBEREG %in% c(
      "Southeast (AL, AR, FL, GA, KY, LA, MS, NC, SC, TN, VA, WV)", 
      "Southwest (AZ, NM, OK, TX)"), 1, 0),
    WEST = ifelse(OBEREG %in% c(
      "Far West (AK, CA, HI, NV, OR, WA)", 
      "Rocky Mountains (CO, ID, MT, UT, WY)"), 1, 0),

    # CARNEGIE CLASSIFICATION
    SZ_MEDIUM = ifelse(CARNEGIESIZE %in% c("Medium"), 1, 0),
    SZ_LARGE = ifelse(CARNEGIESIZE %in% c("Large", "Very Large"), 1, 0),
    RSCH_INSTS = ifelse(CARNEGIERSCH == "Research Colleges and Universities", 1, 0),
    RSCH_HIGH = ifelse(CARNEGIERSCH %in% c("Research 1: Very High Spending and Doctorate Production", "Research 2: High Spending and Doctorate Production"), 1, 0),
    ## leave out low
    MED_HI_EARNS = ifelse(CARNEGIESAEC %in% c("Lower Access, Medium Earnings", "Lower Access, Higher Earnings", "Higher Access, Medium Earnings", "Opportunity Colleges and Universities – Higher Access, Higher Earnings"), 1, 0), 
    ## APM: Program Mixes
    MIXED = ifelse(CARNEGIEAPM == "Mixed", 1, 0),
    PROFESSIONS = ifelse(CARNEGIEAPM == "Professions-focused", 1, 0),
    HEALTH = ifelse(CARNEGIEAPM %in% c(
                                    "Special Focus: Nursing",
                                    "Special Focus: Other Health Professions",
                                    "Special Focus: Medical Schools and Centers"
                                  ), 1, 0),
    ARTS = ifelse(CARNEGIEAPM %in% c(
                                    "Special Focus: Arts and Sciences",
                                    "Special Focus: Arts, Music, and Design"
                                  ), 1, 0),
    BUSINESS = ifelse(CARNEGIEAPM %in% c(
                                    "Special Focus: Business",
                                    "Special Focus: Applied and Career Studies"
                                  ), 1, 0),
    STEM = ifelse(CARNEGIEAPM %in% c("Special Focus: Technology, Engineering, and Sciences"), 1, 0)
    # leave out CARNEGIEAPM == "Special Focus: Theological Studies"
  ) %>%
  # remove the original categorical variables
  select(-CONTROL, -APPFEEUW, -TUITPL1, -TUITPL3, -ADMCON7, -CARNEGIESIZE, -CARNEGIERSCH, -CARNEGIESAEC, -ALLONCAM, -CARNEGIEAPM, -HBCU, -CALSYS, -HOSPITAL, -ASSOC1, -BOARD, -LOCALE, -HDEGOFR1, -OPENPUBL, -OBEREG, -RELAFFIL, -ROOM, -SLO6, -SLOA, -STUSRV2, -STUSRV3)

cat("Number of columns after 1st cleaning =", ncol(private1), "\n")
```

## 2nd Cleaning

Check NAs, we don't want to use variables that have a lot of missing rows (over 5%). Variables like Men, Women we will only keep Men. We will also remove other tuition data like out-of-state tuition, published tuition, and other uninterested variables.

```{r echo=FALSE}
data.frame(
  column = names(private1),
  missing = colSums(is.na(private1)),
  perc_missing = round(colSums(is.na(private1)) / nrow(private1) * 100, 1)
)

# remove cols
private2 <- private1 %>%
  # remove variables with lots of NAs
  select(-DISABPCT, 
         -ADMCON1, 
         -ADMCON2, 
         -DVADM01, 
         -DVADM04, 
         -ADMSSN, 
         -ENRLT, 
         -ACTPCT, -SATPCT, -SATVR25, -SATVR50, -SATVR75, -SATMT25, -SATMT50, -SATMT75, -ACTCM25, -ACTCM50, -ACTCM75, -ACTEN25, -ACTEN50, -ACTEN75, -ACTMT25, -ACTMT50, -ACTMT75, 
         -PROPBOR, -PROPBOR_nr, 
         -AVEDEBT, -AVEDEBT_nr, 
         -CLASIZUND20, -CLASIZ2049, -CLASIZOVE50, 
         -APPLFEEU, -RMBRDAMT, -COAISGS2, -NPIST2, -COAIST42, -EFGNDRUA, -EFGNDRKN, -TRRTTOT, -ADMCON7_REQUIRED) %>%
  # remove variables that is 100% correlated with another
  select(
    -EFFTWOM,  # Total women (other: men)
    -EFUNKNT,  # Race/ethnicity unknown total (other: Specifics like Asian, Black)
    -GRRTW,    # Graduation rate, women (other: men),
    -RMUNKNWP, # Percent of first-time undergraduates - residence unknown
    -EF2MORT,  # Two or more races total
    -EFTOTAL,  # Grand total (We already have Full time total and total enrollment)
    -RMOUSTTP  # Percent of first-time undergraduates - out-of-state (We have in-state already)
    ) %>%
  # remove variables directly part of Tuition or other Tuitions or not related
  select(-INSTNM, -CHG2AT3, -CHG3AT3, -CIPCODE2, -CIPCODE3, -CIPCODE4, -CITY, -COUNTYNM, -COUNTYCD, -LONGITUDE, -LATITUDE, -STATE, -ZIP, -WEBADDR, -TUITION3, -SLOB, -OPENPUBL_YES, -WEST) %>%
  # remove uninterested county data
  select(
    -elderly_share_pop,
    -retirement_income,
    -avg_individuals_per_return
  )

cat("Number of columns after 2nd cleaning = ", ncol(private2), "\n")
```

## 3rd Cleaning

We will now do a third time cleaning:

- Remove any non-numerical variables (to remove extra categorical variables we're not interested in. If the number of columns decrease, should check what extra columns got removed). In this case, we removed ATHASSOC (Member of National Athletic Association) because we have one for NCAA already, removed CARNEGIEIC (Institutional Classification) because we have program mix and this variable is over-complicated.

- Remove any row with NAs

```{r echo=FALSE}
private3 <- private2 %>%
  dplyr::select(UNITID, where(is.numeric)) %>%
  na.omit()

cat("Number of rows after 3rd cleaning = ", nrow(private3), "\n")
cat("Number of columns after 3rd cleaning = ", ncol(private3), "\n")
```

# 2. Data Transformation

## Response Variable: TUITION2

Tuition price is almost always right-skewed and usually need log-transformation but we will check just in case. Here's the distribution of the original TUITION2.


```{r echo=FALSE}
# Check distribution
hist(private3$TUITION2,
     main = "Distribution of TUITION2",
     col = "steelblue")
```

Next, we check the AIC of each model with original TUITION2, log-transformed TUITION2, and sqrt-transformed TUTION2. AIC is the Akaike information criterion, a mathematical method for evaluating how well a model fits the data it was generated from.

```{r}
m1 <- lm(TUITION2 ~ . - UNITID, data = private3)
m2 <- lm(log(TUITION2) ~ . - UNITID, data = private3)
m3 <- lm(sqrt(TUITION2) ~ . - UNITID, data = private3)

AIC(m1, m2, m3)
```

AIC of the model with log-transformed TUITION2 is much lower, so we should use log-transformation. Next, we check Residuals vs Fitted plot for variance behavior.

```{r echo=FALSE}
par(mfrow = c(1,3))

plot(m1$fitted.values, m1$residuals,
     xlab = "Fitted TUITION2", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

plot(m2$fitted.values, m1$residuals,
     xlab = "Fitted log(TUITION2)", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

plot(m3$fitted.values, m1$residuals,
     xlab = "Fitted sqrt(TUITION2)", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)
```

The residual plot with the **original TUITION2** looks fairly random but:

- Slightly wider spread at higher fitted values

- More extreme negative residuals

- Mild heteroskedasticity (variance increases with fitted values)

The one with **log-transformation**, although has tighter clouds, actually has better variance:

- Spread looks more balanced across fitted values

- Extremes are more symmetric

- Variance looks more stable

The one with **square root transformation ** is similar to the original:

- Some widening of spread at higher fitted values

- Slight asymmetry in residuals

So even though the model with the original TUITION2 gives the highest $R^2$, we should choose **log-transformation**. Thus, we add log(TUITION2) to our dataframe and remove TUITION2 column.

```{r}
private4 <- private3 %>%
  mutate(log_TUITION2 = log(TUITION2)) %>%
  select(-TUITION2)
```


## Numeric variables vs log_TUITION2

We will now check numerical variables in pair plots with log_TUITION2 to see if we need to perform transformations.

```{r echo=FALSE}
# Institution Profile & Resources
pairs(private4 %>% dplyr::select(log_TUITION2, income_per_capita, gdp_per_capita, popestimate, STUFACR),
      main = "Matrix Plot: Institution Profile & Resources vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("violet", alpha.f = 0.4))

# Enrollment and student composition
pairs(private4 %>% dplyr::select(log_TUITION2, EFAIANT, EFASIAT, EFBKAAT, EFFT, EFFTMEN),
      main = "Matrix Plot: Enrollment Variables 1 vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("steelblue", alpha.f = 0.4))

pairs(private4 %>% dplyr::select(log_TUITION2, EFHISPT, EFNHPIT, EFNRALT, EFUG1ST, RMINSTTP),
      main = "Matrix Plot: Enrollment Variables 2 vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("steelblue", alpha.f = 0.4))

pairs(private4 %>% dplyr::select(log_TUITION2, EFUGCNT, EFUGTRN, EFWHITT, ENRTOT, RMFRGNCP),
      main = "Matrix Plot: Enrollment Variables 3 vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("steelblue", alpha.f = 0.4))

# Admission and selectivity
pairs(private4 %>% dplyr::select(log_TUITION2, GBA4RTT, GBA6RTT, GRRTM, PGGRRTT),
      main = "Matrix Plot: Admissions & Outcomes vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("darkgreen", alpha.f = 0.4))

# Financial aid and cost variables
pairs(private4 %>% dplyr::select(log_TUITION2, AGRNT_N, AGRNT_T, UAGRNTP, UDGPGRNTN, UFLOANN),
      main = "Matrix Plot: Financial Aid Variables vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("firebrick", alpha.f = 0.4))
```

Most of our numerical variables have extreme right skewness. Thus we will:

- Log-transform log(x + 1) on counts or total dollar amounts (Enrollment & demographic counts, grants, aid and economic resources) since they are heavily right-skewed

- Square-root transformation sqrt(x) on ratios like STUFACR since it has moderate skewness

- No transformation on variables like graduation rates (GBA4RTT, GBA6RTT, GRRTM, PGGRRTT) and certain percentages (UAGRNTP, RMINSTTP) since they appear relatively normally distributed or well-spread

```{r}
private5 <- private4 %>% mutate(
  # Log transformations for heavily right-skewed counts and financial totals
  ## Enrollment & Demographic Counts
  log_EFAIANT    = log(EFAIANT + 1),
  log_EFASIAT    = log(EFASIAT + 1),
  log_EFBKAAT    = log(EFBKAAT + 1),
  log_EFFT       = log(EFFT + 1),
  log_EFFTMEN    = log(EFFTMEN + 1),
  log_EFHISPT    = log(EFHISPT + 1),
  log_EFNHPIT    = log(EFNHPIT + 1),
  log_EFNRALT    = log(EFNRALT + 1),
  log_EFUG1ST    = log(EFUG1ST + 1),
  log_EFUGCNT    = log(EFUGCNT + 1),
  log_EFUGTRN    = log(EFUGTRN + 1),
  log_EFWHITT    = log(EFWHITT + 1),
  log_ENRTOT     = log(ENRTOT + 1),
  log_RMFRGNCP   = log(RMFRGNCP + 1), # Percentage but heavily right-skewed in plot
  
  ## Financial Aid & Economic Resources
  log_AGRNT_N    = log(AGRNT_N + 1),
  log_AGRNT_T    = log(AGRNT_T + 1),
  log_UDGPGRNTN  = log(UDGPGRNTN + 1),
  log_UFLOANN    = log(UFLOANN + 1),
  log_income_per_capita = log(income_per_capita + 1),
  log_gdp_per_capita    = log(gdp_per_capita + 1),
  log_popestimate       = log(popestimate + 1),
  
  # Square Root transformation for moderate skew in ratios
  sqrt_STUFACR   = sqrt(STUFACR)
) %>%
# Remove original columns to avoid multicollinearity with transformed versions
select(-EFAIANT, -EFASIAT, -EFBKAAT, -EFFT, -EFFTMEN, -EFHISPT, -EFNHPIT, 
       -EFNRALT, -EFUG1ST, -EFUGCNT, -EFUGTRN, -EFWHITT, -ENRTOT, -RMFRGNCP,
       -AGRNT_N, -AGRNT_T, -UDGPGRNTN, -UFLOANN, -income_per_capita, 
       -gdp_per_capita, -popestimate, -STUFACR)
```

The transformed numerical variables pair plots look much better than that of the original numerical variables.

```{r echo = FALSE}
# Institution Profile & Resources
pairs(private5 %>% dplyr::select(log_TUITION2, log_income_per_capita, log_gdp_per_capita, log_popestimate, sqrt_STUFACR),
      main = "Matrix Plot: Institution Profile & Resources vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("violet", alpha.f = 0.4))

# Enrollment and student composition
pairs(private5 %>% dplyr::select(log_TUITION2, log_EFAIANT, log_EFASIAT, log_EFBKAAT, log_EFFT, log_EFFTMEN),
      main = "Matrix Plot: Enrollment Variables 1 vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("steelblue", alpha.f = 0.4))

pairs(private5 %>% dplyr::select(log_TUITION2, log_EFHISPT, log_EFNHPIT, log_EFNRALT, log_EFUG1ST),
      main = "Matrix Plot: Enrollment Variables 2 vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("steelblue", alpha.f = 0.4))

pairs(private5 %>% dplyr::select(log_TUITION2, log_EFUGCNT, log_EFUGTRN, log_EFWHITT, log_ENRTOT, log_RMFRGNCP),
      main = "Matrix Plot: Enrollment Variables 3 vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("steelblue", alpha.f = 0.4))

# Admission and selectivity
pairs(private5 %>% dplyr::select(log_TUITION2, GBA4RTT, GBA6RTT, GRRTM, PGGRRTT),
      main = "Matrix Plot: Admissions & Outcomes vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("darkgreen", alpha.f = 0.4))

# Financial aid and cost variables
pairs(private5 %>% dplyr::select(log_TUITION2, log_AGRNT_N, log_AGRNT_T, UAGRNTP, log_UDGPGRNTN, log_UFLOANN),
      main = "Matrix Plot: Financial Aid Variables vs In-State Tuition",
      pch  = 16, cex = 0.3,
      col  = adjustcolor("firebrick", alpha.f = 0.4))
```

```{r echo = FALSE}
m <- lm(log_TUITION2 ~ . - UNITID, data = private5)
cat("AIC prior to transforming numerical values =", AIC(m2), "\n")
cat("AIC of after transformations =", AIC(m), "\n")
```

The decrease in AIC also confirmed that the transformations were appropriate.

# 3. Best Subset Selection

## Group Variables into Categories

We have too many predictors to run one model. So we split into groups as following (Note that this grouping is strictly for best subset and grouping in the R Shiny App should follow the [code book](https://docs.google.com/spreadsheets/d/15nn2SjonRKGXaUnBLJxZs6E04GZDL0sNgCyN5-q8Vig/edit?gid=1289751215#gid=1289751215)).

- Group 1: Enrollment & Demographics

- Group 2: Outcomes & Institutional Characteristics

- Group 3: Financial Aid & Grant

- Group 4: Institutional / Structural Indicators

- Group 5: Geography

- Group 6: Program Mix

```{r echo=FALSE}
group1 <- c(
  "log_ENRTOT",   
  "log_EFUGCNT",
  "log_EFUG1ST",
  "log_EFUGTRN",
  "log_EFFT",
  "log_EFFTMEN",
  "log_EFWHITT",
  "log_EFAIANT",
  "log_EFASIAT",
  "log_EFBKAAT",
  "log_EFHISPT",
  "log_EFNHPIT",
  "log_EFNRALT",
  "log_RMFRGNCP",
  "RMINSTTP"
)

group2 <- c(
  "GBA4RTT", 
  "GBA6RTT", 
  "GRRTM", 
  "PGGRRTT", 
  "sqrt_STUFACR", 
  "SZ_MEDIUM", 
  "SZ_LARGE", 
  "RSCH_INSTS",
  "RSCH_HIGH",
  "MED_HI_EARNS",
  "HDEOFR_DOC",
  "HDEOFR_MAS"
)

group3 <- c(
  "UAGRNTP", 
  "log_AGRNT_N", 
  "log_AGRNT_T", 
  "log_UDGPGRNTN", 
  "log_UFLOANN",
  "APPFEEUW_YES",
  "TUITPL1_YES",
  "TUITPL3_YES"
)

group4 <- c(
  "FORPROFIT",
  "HBCU_YES",
  "HOSPITAL_YES",
  "RELAFFIL_NO",
  "SEMESTER",
  "ALLONCAM_YES",
  "ASSOC1_YES",
  "BOARD_YES",
  "ROOM_YES",
  "SLO6_YES",
  "SLOA_YES",
  "STUSRV2_YES",
  "STUSRV3_YES"
)

group5 <- c(
  "LO_SUBURB",
  "LO_TOWN",
  "LO_CITY",
  
  "NORTHEAST",
  "MIDWEST",
  "SOUTH",
  
  "log_popestimate",
  "log_income_per_capita",
  "log_gdp_per_capita"
)

group6 <- c(
  "MIXED",
  "PROFESSIONS",
  "HEALTH",
  "ARTS",                 
  "BUSINESS",
  "STEM"
)

official_col_count <- ncol(private5) - 2 # 1 for unitid, 1 for log_TUITION2
grouping_count <- sum(length(group1), length(group2),length(group3),length(group4),length(group5),length(group6))

cat("Manual groupings count matches nubmer of columns in private5:", official_col_count - grouping_count == 0, "\n")
```

Next, we run best subset on each group and find the best variables from each group:

```{r echo=FALSE}
# function to run best subset
run_best_subset <- function(data, vars, response = "log_TUITION2") {

  formula <- as.formula(
    paste(response, "~", paste(vars, collapse = " + "))
  )

  regsubsets(
    formula,
    data = data,
    nvmax = 15,
    method = "exhaustive"
  )
}

# function to extract best variables
get_best_vars <- function(model) {

  s <- summary(model)

  best <- which.min(s$bic)

  names(coef(model, best))[-1]
}
```

```{r echo=FALSE}
# run best subset on each grouping
bs1 <- run_best_subset(private5, group1)
plot(bs1)
bs2 <- run_best_subset(private5, group2)
plot(bs2)
bs3 <- run_best_subset(private5, group3)
plot(bs3)
bs4 <- run_best_subset(private5, group4)
plot(bs4)
bs5 <- run_best_subset(private5, group5)
plot(bs5)
bs6 <- run_best_subset(private5, group6)
plot(bs6)
```

Thus, the most important variables are in the pool (lower bic, better model):

```{r echo=FALSE}
# get best variables from each model
final_vars <- c(
  get_best_vars(bs1),
  get_best_vars(bs2),
  get_best_vars(bs3),
  get_best_vars(bs4),
  get_best_vars(bs5),
  get_best_vars(bs6)
)

final_vars
```


## Final Best Subset

Now, we run another best subset with the variables we deem important from the above analysis.

```{r echo=FALSE}
final_formula <- as.formula(
  paste(
    "log_TUITION2 ~",
    paste(final_vars, collapse = " + ")
  )
)

best_subsets <- regsubsets(final_formula, data = private5, nvmax = 36)
summary_obj <- summary(best_subsets)
mat <- summary_obj$which          # logical matrix: rows = model size, cols = variables
adjr2 <- summary_obj$adjr2        # adj R² for each model size

# Build a data frame
df_plot <- as.data.frame(mat)
df_plot$adjr2 <- round(adjr2, 2)
df_plot$model <- seq_len(nrow(df_plot))

df_long <- melt(df_plot, id.vars = c("model", "adjr2"),
                variable.name = "Variable", value.name = "Included")

ggplot(df_long, aes(x = Variable, y = factor(adjr2), fill = Included)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  labs(title = "In-State Best Subsets (Transformed)",
       x = NULL, y = "adj R²") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9),
    legend.position = "none",
    panel.grid = element_blank()
  )
```

# Linear Model

We now create a linear model with the selected variables from the best subset. We will include all n-1 levels of a cetgorical variable we transformed into dummies.

```{r echo=FALSE}
final_model <- lm(log_TUITION2 ~ 
                    log_ENRTOT +
                    log_EFUGCNT +
                    log_EFASIAT +
                    log_EFHISPT +
                    log_EFWHITT +
                    log_EFNHPIT +
                    GBA6RTT +
                    GRRTM +
                    RSCH_HIGH + RSCH_INSTS +
                    log_AGRNT_N +
                    log_AGRNT_T +
                    log_UDGPGRNTN +
                    log_UFLOANN +
                    HBCU_YES +
                    ASSOC1_YES +
                    SLO6_YES +
                    MIXED + HEALTH + PROFESSIONS + ARTS + STEM +
                    SZ_MEDIUM + SZ_LARGE
                    - UNITID, data = private5)

summary(final_model)
```

We also generate plots to validate the model.

```{r echo=FALSE}
par(mfrow = c(2,2))
plot(final_model)
```

**Next step**: identifying outliers, validating models, run this for out-of-state tuition, public schools, etc.

